|

### 2024-08-29 reproducibility/manuscript/manuscript.qmd
bugfix with align vs aligned
updating later methods section to have consistent notation

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{equation}
  \label{eq-rna}
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{align}
\end{equation}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)}.
\end{align}
\end{equation}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are, in fact, unknown.

@@

We define the ODE solution at time $\tau_{cg}$ as:
\begin{equation}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed counts as:
\begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa].
For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
\begin{equation}
\label{eq-rna-solution}
\begin{align}
  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
  s(t) &= \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{align}
\end{equation}
These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.

@@

Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
we would like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align}
but exact Bayesian inference is intractable in this model. We use Pyro to
automatically integrate out the local discrete latent variables $k$, which is
defined as the cell and gene transcriptional state (see above), and approximate
the posterior over the remaining latent variables using variational inference
[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
optimization problems that can be solved with off-the-shelf tools. In
variational inference, we maximize a tractable bound (the evidence lower bound
(ELBO)) on the Kullback-Leibler (KL) divergence
$$
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
$$
between a parametric family of tractable probability distributions q and the
intractable true posterior
\begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align}
We approximate our model's posterior distribution $p( \theta, \psi \vert
\tilde{X})$ with a tractable family of probability distributions
$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
location parameter and a learnable low-rank covariance matrix. We solve the
resulting stochastic optimization problem using a version of Automatic
Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
update the variational parameters  using stochastic gradient ascent.  Our
implementation and experiments use generic variational families and Monte Carlo
ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
library for probabilistic machine learning, and a version built into Pyro of the
adaptive stochastic gradient ascent algorithm Adam augmented with gradient
clipping to enhance numerical stability during training.

@@

To benchmark Pyro&thinsp;-Velocity performance in predicting cell fate, we
generated the posterior samples measurement
$x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
s\left(\tau^{(k_{cg})}\right)\right)$ or $x_{cg}=\left(u_{cg}, s_{cg} \right)$,
$t_c$, $\beta_g$, $\gamma_g$ from the same single cell RNA-seq using $N=30$
Monte Carlo samples from the posterior predictive distribution following
$p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
\psi) q_{\phi}(\theta, \psi)$. $x_{cg}$ can be posterior samples of either
denoised gene expression (used for phase portraits and vector field-based
trajectory inference) or raw read counts (used for gene selection). Then we
calculated the posterior samples of RNA velocity as
$v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}$ based on posterior samples
measurement of denoised gene expression $x_{cg}$ and $\beta_{g}$, $\gamma_{g}$.

@@

We prioritize the cell fate markers using two metrics: First, the Pearson
correlation between each gene's posterior mean of the denoised spliced
expression and posterior mean of the shared time; Second, the negative mean
absolute errors of each gene's observed spliced, and unspliced read counts with
posterior predictive samples of spliced and unspliced raw read counts, i.e.
$\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})$, where $N$ is the
posterior sample number that is set to $30$, $i$ is the posterior sample index,
and $n_c$ is the cell number. We first select the top $300$ genes with the
highest negative mean absolute errors and then rank the $300$ genes based on the
most positively correlated genes and the least negatively correlated genes. We
use the same strategy for scVelo to rank the markers by the model likelihood to
get the top 300 genes and then prioritize these genes by Pearson's correlation
between scVelo latent time and normalized expression.

@@

Pyro&thinsp;-Velocity evaluates cell fate uncertainty using four statistics. The
first is the uncertainty of shared time, which is evaluated using the standard
deviation of posterior samples of $t_c$ for each cell. The second statistic is
the angular standard deviation [@Berens2009-fu] across posterior samples of the
velocity-based vector field. The value of this statistic was mapped linearly to
the range $[0, 360]$ for interpretability. The third statistic is the Rayleigh
test [@The-Astropy-Collaboration2022-aw] for posterior samples of the
velocity-based vector fields for each cell, whose p-values have been corrected
for multiple tests using Benjamini-Hochberg FDR methods; if FDR is less than
$0.05$, the cell is considered to have one statistically significant future
direction. The second and third circular statistics were implemented using the
Astropy package (`v5.1`) [@The-Astropy-Collaboration2022-aw]. The last statistic
is state uncertainty which evaluates the mean Euclidean distance between
posterior mean and posterior samples of the raw read count prediction for each
cell
\begin{align*}
\frac{1}{N} \sum_{i} \sqrt{\sum_{g} \left(x_{icg} - \frac{1}{N}\sum_i x_{icg}\right)^2}.
\end{align*}

@@

Pyro&thinsp;-Velocity provides two model formulations as described above.
Users can compare their predictive performance to decide which of the two models
to use for a given dataset. Namely, the mean absolute errors of posterior
samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
observation $\tilde{X}_{cg}$ per gene and cell combinations
\begin{align*}
\frac{1}{N_i N_c N_g} \sum_{i=1}^{N_i} \sum_{c=1}^{N_c} \sum_{g=1}^{N_g} 
  \vert x_{icg} - \tilde{X}_{cg} \vert,
\end{align*}
where $i$ is the posterior sample index, $N_i$ is the posterior sample number,
$N_c$ is the cell number, $N_g$ is the gene number. Then it is possible to
evaluate the cell fate inference performance of the models based on two metrics:

>>>>

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
ordinary differential equation (ODEs) proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{equation}
  \label{eq-rna}
\begin{aligned}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{aligned}
\end{equation}
where $u(t), s(t)$ are the unspliced and spliced expression levels of a gene at time $t$ under a transcription rate $\alpha(t)$ with possible temporal dependence, splicing rate $\beta$, and degradation rate $\gamma$. We specify this model to a setting that depends on cell $c$ and gene $g$ as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{aligned}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-du-cg} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-ds-cg} .
\end{aligned}
\end{equation}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u_{cg}(t), s_{cg}(t) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene.
We restrict attention to piecewise-constant $\alpha_{cg}(t)$ to capture gene-specific activation and repression. We take special care to model a gene- and cell-specific switching time that marks a single transition from activation to repression by introducing a Bernoulli variable $k_{cg}$ to model unknown transcription state. We assume our cell-by-gene data-matrix arrive as observations of Poisson-counts related to the solution of the above ODEs at unknown times, which is modeled as a relationship between an unknown latent time shared across each cell, $t_c$, and unknown gene-specific time-offsets $t_{0,g}$ where all read counts for a single cell occurred at an unknown, but shared latent time $t_c$. These relative times are also used to parametrize the Bernoulli process for $k_{cg}$. 
Importantly, we recognize that the initial conditions are, in fact, unknown.

@@

We define the ODE solution at time $\tau_{cg}$ as:
\begin{equation}
  \label{eq-ODE-solve}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed counts as:
\begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big)   \label{eq-u-hat-obs} \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big) \label{eq-s-hat-obs}
\end{align}
Here, we use ${u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}$ to denote the observed unspliced and spliced counts
for cell $c$ and gene $g$. We use $\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}$ to
denote our generative model's prediction of these unspliced and spliced expression levels.
The generative process for modeling these observed read counts given denoised gene transcript expression level $\hat{u}_{cg}, \hat{s}_{cg}$ considers the expected number of observed reads for a given gene in a given cell as the number of transcripts times the ratio of the cell's total reads to total transcripts.

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the assumption that the transcription rate $\alpha(t)$ is piece-wise constant, and is derived by the authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa].
For brevity, we present the solution under constant transcription rate $\alpha(t) = \alpha_0$, and note that this can be combined to assemble the solution in the piece-wise constant case.
First, let $C = u_0 - \frac{\alpha}{\beta}$. Then, the solution to the ODE in Eq. \ref{eq-rna} is:
\begin{equation}
\label{eq-rna-solution}
\begin{aligned}
  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
  s(t) &= \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{aligned}
\end{equation}
These solutions are the essential building block used to compute all cell-gene-specific ODE solutions for the unspliced and spliced expression levels in the generative model.

@@

Given observations $\tilde{X}_{cg} = \left( u_{cg}^{\text{(obs)}}, s_{cg}^{\text{(obs)}} \right)$ over cells $1 \dots C$ and genes $1 \dots G$,
we would like to compute the posterior distribution over the random variables $\theta := \left( \theta_1, \ \dots, \theta_C \right)$ and $\psi := \left( \psi_1, \ \dots, \psi_G \right)$, where
\begin{align}
\theta_c &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi_g &= \left( {t_{0,g}^{(0)}}, \Delta \text{switching}_{g}, 
          \alpha^{(0)}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align}
but exact Bayesian inference is intractable in this model. We use Pyro to
automatically integrate out the local discrete latent variables $k$, which is
defined as the cell and gene transcriptional state (see above), and approximate
the posterior over the remaining latent variables using variational inference
[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
optimization problems that can be solved with off-the-shelf tools. In
variational inference, we maximize a tractable bound (the evidence lower bound
(ELBO)) on the Kullback-Leibler (KL) divergence
$$
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
$$
between a parametric family of tractable probability distributions q and the
intractable true posterior
\begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align}
We approximate our model's posterior distribution $p( \theta, \psi \vert
\tilde{X})$ with a tractable family of probability distributions
$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
location parameter and a learnable low-rank covariance matrix. We solve the
resulting stochastic optimization problem using a version of Automatic
Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
update the variational parameters  using stochastic gradient ascent.  Our
implementation and experiments use generic variational families and Monte Carlo
ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
library for probabilistic machine learning, and a version built into Pyro of the
adaptive stochastic gradient ascent algorithm Adam augmented with gradient
clipping to enhance numerical stability during training.

@@

To benchmark Pyro&thinsp;-Velocity performance in predicting cell fate, we
generated posterior samples of quantitites of interest in the model.
In particular, we focus on posterior samples of $i)$ spliced and unspliced gene expression across all cells and genes, defined in Eq. \ref{eq-ODE-solve} and denoted by
$X_{cg}=\left(u_{cg}, s_{cg} \right)$ (used for phase portraits and vector-field-based
trajectory inference),
$ii)$ read counts, defined in Eqs. \ref{eq-u-hat-obs} and \ref{eq-s-hat-obs} and denoted by
$\hat{X}^{\text{(obs)}}_{cg}=\left(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg} \right)$
(used for gene selection),
and
$iii)$ RNA velocity, defined in Eq. \ref{eq-ds-cg} as the time derivative of the spliced expression, denoted by
$v_{cg}=\frac{ds_{cg}}{d t}(X_{cg})$.
We generate 30 Monte Carlo samples from the posterior predictive distribution
$X, \hat{X}^{(\text{obs})}, v \sim p(X, \hat{X}^{(\text{obs})}, v \vert \theta, \psi) p(\theta, \psi) \approx q_{\phi}(\theta, \psi)$ 
by sampling the posterior distribution of $\theta$, $\psi$ using the trained model.
We use these samples to compute statistics on the posterior predictive distribution of our quantities of interest; these statistics are used to evaluate the model's performance and uncertainty in predicting cell fate.

@@

We prioritize the cell fate markers using two metrics: First, the Pearson
correlation between each gene's posterior mean of the denoised spliced
expression and posterior mean of the shared time; Second, the negative mean
absolute errors of each gene's observed spliced, and unspliced read counts with
posterior predictive samples of spliced and unspliced raw read counts, i.e.
$\mathcal{E}_g = \frac{1}{NC} \sum_{i=1}^N \sum_{c=1}^{C} \vert \tilde{X}_{cg} - \hat{X}_{cg}^{(\text{obs}), (i)} \vert$, where we use $N=30$ Monte Carlo samples. We first select the top $300$ genes with the
highest negative mean absolute errors and then rank the $300$ genes based on the
most positively correlated genes and the least negatively correlated genes. We
use the same strategy for scVelo to rank the markers by the model likelihood to
get the top 300 genes and then prioritize these genes by Pearson's correlation
between scVelo latent time and normalized expression.

@@

Pyro&thinsp;-Velocity evaluates cell fate uncertainty using four statistics. The
first is the uncertainty of shared time, which is evaluated using the standard
deviation of posterior samples of $t_c$ for each cell. The second statistic is
the angular standard deviation [@Berens2009-fu] across posterior samples of the
velocity-based vector field. The value of this statistic was mapped linearly to
the range $[0, 360]$ for interpretability. The third statistic is the Rayleigh
test [@The-Astropy-Collaboration2022-aw] for posterior samples of the
velocity-based vector fields for each cell, whose p-values have been corrected
for multiple tests using Benjamini-Hochberg FDR methods; if FDR is less than
$0.05$, the cell is considered to have one statistically significant future
direction. The second and third circular statistics were implemented using the
Astropy package (`v5.1`) [@The-Astropy-Collaboration2022-aw]. The last statistic
is state uncertainty which evaluates the mean Euclidean distance between
posterior mean and posterior samples of the raw read count prediction for each
cell
\begin{align*}
\mathcal{U}_c = \frac{1}{N} \sum_{i=1}^N \sqrt{\sum_{g=1}^G \left(\hat{X}^{\text{(obs)},(i)}_{cg} - \frac{1}{N}\sum_{i=1}^N \hat{X}^{\text{(obs)},(i)}_{cg}\right)^2}
\end{align*}
with $N=30$ Monte Carlo samples.

@@

Pyro&thinsp;-Velocity provides two model formulations as described above.
Users can compare their predictive performance to decide which of the two models
to use for a given dataset. Namely, the mean absolute errors of posterior
samples prediction of spliced and unspliced raw read counts $x_{icg}$ from its
observation $\tilde{X}_{cg}$ per gene and cell combinations
\begin{align*}
\frac{1}{N C G} \sum_{i=1}^{N} \sum_{c=1}^{C} \sum_{g=1}^{G} 
  \vert X^{\text{(obs), (i)}}_{cg} - \tilde{X}_{cg} \vert,
\end{align*}
where $i$ is the posterior sample index, $N$ is the posterior sample number,
$C$ is the cell number, $G$ is the gene number. Then it is possible to
evaluate the cell fate inference performance of the models based on two metrics:
